---
title: "Analysis of single-cell RNA-seq data using flashier"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Analysis of single-cell RNA-seq data using flashier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The aim of this vignette is show how flashier can be used to analyze
single-cell RNA-seq data. This vignette is modeled after the
[fastTopics vignette on analyzing single-cell data][fasttopics-vignette].

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 6, warning = FALSE,
                      results = "hold", fig.align = "center", dpi = 120)
```

We begin by loading the required packages. We also set the seed so that results
can be fully reproduced.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(reshape)
library(flashier)
library(fastTopics)
library(ggplot2)
library(cowplot)
set.seed(3)
```

Preparing the single-cell data for flashier
-------------------------------------------

The single-cell RNA-seq data (we use the `pbmc_facs` data from the
`fastTopics` package) are unique molecular identifier (UMI) counts
stored as an $n \times p$ sparse matrix, where $n$ is the number of
cells and $p$ is the number of genes:

```{r load-data}
data("pbmc_facs")
counts <- pbmc_facs$counts
dim(counts)
```

(Note that other R packages, for example Seurat, use the convention
that rows are genes and columns are cells.)

Since the flashier model, like other linear-model-based methods (e.g.,
principal components analysis), was not designed for count data, it is
recommended to first transform the data in a way that makes them more
suitable for flashier. A widely used approach is to divide the counts
by a "size factor," add a "pseudocount," then take the log. We call
this transformation the "shifted logarithm," following
[this paper][ahlmann-eltze-huber-2023], and we refer to the
transformed counts as "shifted log counts".
  
In practice, the shifted log counts are computed as follows: we first
divide the counts by $\alpha s_i$, where $s_i$ is the size factor for
cell $i$, and $\alpha$ is the pseudocount. Done this way, the shifted
log counts maintain sparsity in the data; that is, if the original
count is zero, then the shifted log count is also zero.
  
More formally, the transformed counts $y_{ij}$ obtained from the
original counts $x_{ij}$ are
$$
y_{ij} = \log\bigg(1 + \frac{x_{ij}}{\alpha s_i}\bigg).
$$
In our analysis below, we use library-size normalization;
that is, we set
$$
s_i =
\frac{\text{total count for cell} \; i}
     {\text{average across all cells}} =
\frac{\sum_{j=1}^p x_{ij}}
     {\frac{1}{n}\sum_{i=1}^n \sum_{j=1}^p x_{ij}},
$$
and we set the pseudocount $\alpha$ to 1.

```{r shifted-log-counts}
a <- 1
size_factors <- rowSums(counts)
size_factors <- size_factors / mean(size_factors)
shifted_log_counts <- log1p(counts / (a * size_factors))
```

Note that, since the counts are sparse, we could have performed this
transformation more efficiently using, for example, the "mapSparse"
function from [the MatrixExtra package][matrixextra].

Variance regularization
-----------------------

Before we run flashier, one issue we need to confront is that flashier
may automatically estimate the variances to be too small, which can
especially be an issue for genes with low expression. We can avoid
this issue by setting a sensible lower bound on the variance
estimates.

<!-- When running flashier, we need to place some regularization on the -->
<!-- variance estimates. Without regularization, it is possible for some -->
<!-- variance estimates to be very small, which can lead to overfitting. -->

We use the following rule of thumb: estimate the standard deviation of
the transformed data for a Poisson random variable with rate $\mu = 1
/ n$, where $n$ is the number of samples. This standard deviation
corresponds to a gene for which we would expect to observe a single
count across all $n$ cells, so it can serve as a reasonable lower
bound to prevent variance estimates from getting too small.

```{r variance-lower-bound}
n  <- nrow(counts)
x  <- rpois(1e7, 1/n)
s1 <- sd(log(x + 1))
```

Fit a flashier model
--------------------

Now we can call flash to fit a flashier model to the transformed
data:

```{r flash}
fit <- flash(shifted_log_counts, 
             ebnm_fn = ebnm_point_exponential,
             var_type = 2, 
             greedy_Kmax = 8, 
             S = s1)
```

A few notes about this flash call:

+ `ebnm_fn = ebnm_point_exponential` forces both the ${\bf L}$ and
  ${\bf F}$ matrices in flashier to be non-negative, so this call will
  generate a non-negative matrix factorization (NMF) of the
  (transformed) counts matrix. Other types of matrix factorizations
  can be produced with different choices of the `ebnm_fn` argument;
  this is explained in detail in the "Introduction to flashier"
  vignette.

+ `var_type = 2` means that we estimate column-wise (here, gene-wise)
  variances; that is, we estimate a different variance for each column
  (gene).

+ `greedy_Kmax = 8` forces flashier to fit no more than 8 factors. **In
  practice we recommend using a larger value,** but to keep the example
  simple and short we set it to 8 here.

Visualizing the memberships
---------------------------

<!-- Each cell in the single-cell data set is represented as a linear
combination of the $K$ factors. -->

The estimated "membership levels", or "memberships", are stored in the
${\bf L}$ matrix. For example:

```{r membership-example}
cell_ids <- c("GATATATGTCAGTG-1-b_cells",
              "GACAGTACCTGTGA-1-memory_t",
              "TGAAGCACACAGCT-1-b_cells")
round(fit$L_pm[cell_ids, ], digits = 3)
```

The first and third cells mainly have membership in factors 1 and 6,
whereas the second cell mostly has membership in factors 1 and 5.
Most of the memberships are zero.

We can use a "structure plot" to visualize the relationship between
the memberships and the cell labels, and possibly get some clues about
the biological meaning of the factors. The cell labels are the
"sorted" cell subpopulations:

```{r cell-samples}
summary(pbmc_facs$samples$subpop)
```
   
A structure plot is a stacked bar plot in which each factor is
represented as a bar of a different color, and the bar heights are
given by the membership levels. For a more effective visualization, we
recommend extracting a rescaled version of the memberships using the
`ldf` function:

```{r ldf}
res <- ldf(fit, type = "i")
```

(Note that we can multiply the $k$th column of ${\bf F}$ by any constant
scalar $c > 0$ while also multiplying the $k$th column of
${\bf L}$ by $1/c$ without changing the low-rank representation, ${\bf
L}{\bf F}^{\top}$.)

Now we can visualize the rescaled memberships in a structure plot. To
do this, we create a simple wrapper function that converts the
flashier results into an object that is usable by the "structure_plot"
function in fastTopics:

```{r structure-plot-1, fig.height=2, fig.width=7, results="hide", message=FALSE}
flashier_structure_plot <- function (ldf, topics, grouping) {
  class(ldf) <- c("multinom_topic_model_fit", "list")
  p <- structure_plot(ldf, topics = topics, grouping = grouping, gap = 20) +
    labs(y = "membership", color = "factor", fill = "factor")
  return(p)
}
flashier_structure_plot(res, c(2:8, 1), pbmc_facs$samples$subpop)
```

The first factor is present in all cell types, and is likely capturing
a "baseline" level of expression throughout. To focus on differences
between cell types, let's remove this first factor from our
structure plot:

```{r structure-plot-2, fig.height=2, fig.width=7, results="hide", message=FALSE}
flashier_structure_plot(res, 2:8, pbmc_facs$samples$subpop)
```
  
It is quite clear from this structure plot that factors 2 through 6
are capturing, respectively, natural killer (NK) cells, CD14+ cells,
CD34+ cells, T cells and B cells. There is also more subtle structure
captured by the memberships, such as the subset of T cells with mixed
memberships (factors 2 and 5, colored blue and orange in the plot),
which suggest a function that is intermediate between that of the
"pure" T and the "pure" NK cells. It is also interesting to note the
subset of CD34+ cells that are primarily loaded on factors 3 or 6
(yellow and green); some of these cells may be mislabeled.

Note that factors 7 and 8 are present in very few cells and probably
offer little insight:

```{r inspect-loadings}
colSums(res$L > 0.01)
```

<!-- Others may have a hybrid function. -->
  
We can also visualize the memberships with a heatmap:

```{r heatmap, fig.height=3.5, fig.width=3.25, message=FALSE}
flashier_heatmap <- function (L, grouping, gap = 0) {
  n <- nrow(L)
  k <- ncol(L)
  colnames(L) <- paste0("k",1:k)
  rows <- order(grouping)
  pdat <- melt(as.data.frame(L[rows,]),variable_name = "factor")
  pdat <- cbind(pdat,grouping = grouping[rows])
  pdat <- cbind(pdat,pos = 1:n + gap*as.numeric(pdat$grouping))
  return(ggplot(pdat,aes(x = factor,y = pos,fill = value)) +
    geom_tile(width = 0.8) +
    scale_fill_gradient(low = "white",high = "darkblue") +
    scale_y_continuous(breaks = tapply(pdat$pos,pdat$grouping,mean)) +
    labs(y = "",fill = "membership") +
	theme_cowplot(font_size = 10))
}
flashier_heatmap(res$L,pbmc_facs$samples$subpop,gap = 50)
```

Interpreting the factors by looking at changes in expression
------------------------------------------------------------

Above, we used the provided cell labels to interpret five of the
factors as capturing distinct cell types (B cells, T cells,
etc). However, one might not always have cell labels, or the existing
cell labels may not be informative for the factors. We can also look
to the ${\bf F}$ matrix for interpreting the factors.

The ${\bf F}$ matrix contains the estimated changes in expression,
specifically changes in expression on the log scale (multiplicative
changes). (The exception to this is the first factor, which in this
example is capturing a "baseline" level of expression.) Intuitively,
genes that were estimated to have the largest increases in expression
might give the most helpful clues about the underlying biology. For
example, many of the genes with the largest expression increases in
factor 6 are also genes characteristic of B cells:

```{r top-genes-factor-6}
F <- with(res,F %*% diag(D))
rownames(F) <- pbmc_facs$genes$symbol
head(sort(F[,6],decreasing = TRUE),n = 16)
```

Let's now apply this simple rule-of-thumb and examine the top 4 genes
(by expression increase) for factors 2 through 6:

```{r top-genes-all, fig.height=4, fig.width=2.5, message=FALSE}
plot_top_genes <- function (F, n = 4) {
  F <- as.data.frame(F)
  k <- ncol(F)
  rows <- lapply(F,function (x) head(order(x,decreasing = TRUE),n = n))
  rows <- rev(unique(unlist(rows)))
  n    <- length(rows)
  F    <- F[rows,]
  pdat <- melt(F,variable_name = "factor")
  pdat <- cbind(pdat,pos = 1:n)
  return(ggplot(pdat,aes(x = factor,y = pos,fill = value)) +
    geom_tile(width = 0.8,color = "white",linewidth = 0.5) +
    scale_fill_gradient(low = "white",high = "darkred") +
    scale_y_continuous(breaks = 1:n,labels = rownames(F)) +
    labs(y = "") +
    theme_cowplot(font_size = 10))
}
colnames(F) <- paste0("k",1:8)
plot_top_genes(F[,2:6])
```

Indeed, this approach reveals genes characteristic of the other cell
types (e.g., GNLY for NK cells, S100A9 for CD14+ cells), although not
always; for example, some genes show large expression increases more
than one factor.

We caution that while this simple rule-of-thumb—annotating the factors
by the genes with the largest increases in expression—worked well
in this example, it may not always be helpful.

Session info
------------

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```

[matrixextra]: https://github.com/david-cortes/MatrixExtra
[zheng-2017]: https://doi.org/10.1038/ncomms14049
[fasttopics-vignette]: https://stephenslab.github.io/fastTopics/articles/single_cell_rnaseq_basic.html
[ahlmann-eltze-huber-2023]: https://doi.org/10.1038/s41592-023-01814-1
