---
title: "Analysis of single-cell RNA-seq data using flashier"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Analysis of single-cell RNA-seq data using flashier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The aim of this vignette is show how flashier can be used to analyze
single-cell RNA-seq data. This vignette is modeled after the
[fastTopics vignette on analyzing single-cell data][fasttopics-vignette].

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 6, warning = FALSE,
                      results = "hold", fig.align = "center", dpi = 120)
```

We begin by loading the packages. We also set the seed so the results
can be fully reproduced.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(flashier)
library(fastTopics)
library(cowplot)
library(ggplot2)
library(ggrepel)
library(pheatmap)
set.seed(3)
```

Preparing the single-cell data for flashier
-------------------------------------------

The single-cell RNA-seq data are unique molecular identifier (UMI)
counts stored as an $n \times m$ sparse matrix, where $n$ is the
number of cells and $m$ is the number of genes.

```{r load-data}
data("pbmc_facs")
counts <- pbmc_facs$counts
dim(counts)
```

Since the flashier model, like other linear-model-based methods (e.g.,
principal components analysis), was not designed for count data, it is
recommended to first transform the data in a way that makes them more
suitable for flashier. A widely used approach is to divide the counts
by a "size factor", add a "pseudocount", then take the log. We call
these transformed counts "shifted log counts".
  
In practice, we modify this approach slightly: we first multiply the
counts by $1/(\alpha s_i)$, where $s_i$ is the size factor for cell
$i$, and $\alpha$ is the pseudocount. This approach has the benefit of
maintaining sparsity in the data; that is, if the original count is
zero, then the transformed value is also zero.
  
More formally, we compute the transformed counts $y_{ij}$ from the
original counts $x_{ij}$ as follows:
$$
y_{ij} = \log\bigg(\frac{x_{ij}}{\alpha s_i} + 1\bigg)
$$
in which
$$
s_i =
\frac{\text{total count for cell} \; i}
     {\text{average across all cells}} =
\frac{\sum_{j=1}^m x_{ij}}
     {\frac{1}{n}\sum_{i=1}^n \sum_{j=1}^m x_{ij}}.
$$
In our analysis below, we set the pseudocount $\alpha$ to 1.

```{r log1p-sparse}
log1p_sparse <- function (counts) {
  out   <- counts
  out@x <- log1p(counts@x)
  return(out)
}
```

```{r}
a <- 1
n             <- nrow(counts)
size_factors  <- rowSums(counts)/(sum(counts)/n)
scaled_counts <- counts/(a*size_factors)
shifted_log_counts <- log1p_sparse(scaled_counts)
```

Variance regularization
-----------------------

When running `flashier`, we need to place some regularization on the
variance estimates. Without regularization, it is possible for some
variance estimates to be very small, and thus lead to overfitting. We
suggest the following rule of thumb: compute the standard deviation of
the transformed data for a Poisson random variable with rate $\mu =
1/n$, where $n$ is the number of samples. ($\mu = 1/n$ is a lower
bound on the rate for each gene). This standard deviation can be
thought of as a lower bound on the variance we expect, and it is used
to prevent our variance estimates from getting too small.

```{r}
mhat = 1/nrow(counts) #estimate of rate
xx = rpois(1e7, mhat) #random poisson
S1 <- sd(log(xx + 1)) #sd of log(X+1)
```

Fit a flashier model
--------------------

Now we fit the `flashier` model to the transformed data. To do this,
we make a call to the function `flash`:

```{r, eval=FALSE}
fit.nn = flash(shifted_log_counts, ebnm_fn = c(ebnm::ebnm_point_exponential, ebnm::ebnm_point_exponential), var_type = 2, greedy_Kmax = 20, S = S1)
```
    
When we call `flash`, we specify a few things. First, we specify the
family of prior distributions for the loading and factor
matrices. Here, we choose the point-exponential distribution for
both. One note on this choice is it forces the entries of the loading
and factor matrices to be non-negative. We also specify a constraint
on the structure of the variance. Here, we set `var_ype = 2`, which
corresponds to assuming column-wise variance. This means we estimate
one variance parameter for each column of the data set. Next, we
specify the maximum number of factors to be added in the factor
matrix. Here, we set `greedy_Kmax = 20`. However, it is possible that
the final model fit will contain fewer factors; `flash` only adds
factors if they help maximize the objective function. Lastly, we
specify `S`, the standard errors. Here, we set `S=S1`, our realistic
lower bound on the variance.

Scaling the membership and factor matrices
------------------------------------------

The loading and factor matrix pair are unique up to a scaling. We may desire to normalize the loading matrix such that for each factor, the maximum loading value is 1. We should also scale the factor matrix accordingly.  
```{r, eval=FALSE}
L.scaled <- t(t(fit.nn$L_pm)/apply(fit.nn$L_pm,2, max))
F.scaled <- t(t(fit.nn$F_pm)*apply(fit.nn$L_pm,2, max))
```
  
These scaled versions are particularly helpful for visualizations.  

Looking at memberships
--------------------
For a flashier fit with $K$ factors, each cell $i$ is represented as a convex combination of these factors. For each cell, we have $K$ factor membership levels -- they are all stored in the $n \times K$ loading matrix. These proportions are learned from the data.  
  
Here are some examples:
```{r, eval=FALSE}
rows <- c("GATATATGTCAGTG-1-b_cells",
          "GACAGTACCTGTGA-1-memory_t",
          "TGAAGCACACAGCT-1-b_cells")
round(fit.nn$L_pm[rows,],digits = 3)
```
    
We can see that the sample `GATATATGTCAGTG-1-b_cells` primarily shows membership in factors 1 and 6, with very small levels of membership in factors 5 and 9. Meanwhile, the sample `GACAGTACCTGTGA-1-memory_t` primarily shows membership in factors 1 and 5, with very small membership in factor 14.  

Interpreting factors using available cell labels
--------------------
In the PBMC data, in addition to gene expression data, we have cell labels. Each cell is assigned to one of five labels (each corresponding to a cell type): B cells, CD14+ monocytes, CD3+ cells, natural killer (NK) cells, and T cells. This additional information can be used to help interpret our factors.  
```{r, eval=FALSE}
samples <- pbmc_facs$samples
summary(samples$subpop)
```
    
We can visualize the loadings using a structure plot. A structure plot is a stacked bar plot in which each factor is represented as a bar of a different color, and the bar heights are membership levels. This plot can help visualize the relationship between cell labels and factor membership levels.  
  
For this structure plot, we visualize the scaled loadings from earlier.

```{r, eval=FALSE}
structure_plot_general = function(Lhat,Fhat,grouping,title=NULL, loadings_order = 'embed', print_plot=FALSE, seed=12345, n_samples = NULL, gap=40, show_legend=TRUE, K = NULL){
  set.seed(seed)

  #if not told to plot all samples, then plot a sub-sample
  if(is.null(n_samples)&all(loadings_order == "embed")){
    n_samples = 2000
  }

  #normalize L such that each factor has a maximum loading value of 1
  #results in an error if all the entries of a column are 0
  Lhat = apply(Lhat,2,function(z){z/max(max(z),1)})
  
  #if not told to plot all factors, then plot the requested subset
  if(!is.null(K)){
    Lhat = Lhat[,1:K]
    Fhat = Fhat[,1:K]
  }
  
  Fhat = matrix(1,nrow=3,ncol=ncol(Lhat))
  
  #add column names to Lhat if it doesn't have column names
  if(is.null(colnames(Lhat))){
    colnames(Lhat) <- paste0("k",1:ncol(Lhat))
  }
  
  #define multinom_topic_model_fit for structure plot function
  fit_list <- list(L = Lhat,F = Fhat)
  class(fit_list) <- c("multinom_topic_model_fit", "list")
  
  #plot
  p <- fastTopics::structure_plot(fit_list,grouping = grouping, loadings_order = loadings_order, n = n_samples, colors = rainbow(ncol(Lhat)), gap = gap,verbose=F) + labs(y = "loading",color = "dim",fill = "dim") + ggtitle(title)
  if(!show_legend){
    p <- p + theme(legend.position="none")
  }
  if(print_plot){
    print(p)
  }
  return(p)
}
```

```{r, eval=FALSE}
structure_plot_general(L.scaled, F.scaled, grouping = samples$subpop, n_samples = nrow(L.scaled))
```
  
We can see that the first factor is somewhat equally present in all cell types. (The first factor is often interpreted as an intercept factor). To focus on the differences between cell types, we can take that factor out of our visualization.

```{r, eval=FALSE}
F0 <- F.scaled
L0 <- L.scaled
F0[,1] <- 0
L0[,1] <- 0
structure_plot_general(L0, F0, grouping = samples$subpop, n_samples = nrow(L0))
```
  
From the structure plot, we can see that NK cells, B cells, CD14+ cells, and T cells have high membership levels in factors 2, 6, 3, and 5, respectively.  
  
We can also visualize the loadings with a heatmap. A heatmap visualizes the matrix of loading values, and a more saturated color corresponds to a higher value. Again, we use the scaled loadings.

```{r, eval=FALSE}
#code for heatmap
loadings_heatmap <- function(fit_L, grouping, sample.names, grouping_col, subjects_subset = NULL){
  if (is.null(subjects_subset) == FALSE){
    anno <- data.frame(subpop = grouping[(subjects_subset)])
    anno_colors <- list(subpop = grouping_col)
  }
  else{
    anno <- data.frame(subpop = grouping)
    anno_colors <- list(subpop = grouping_col)
  }
  if (is.null(colnames(fit_L))){
    colnames(fit_L) <- paste0('k',c(1:ncol(fit_L)))
  }
  if (is.null(rownames(fit_L))){
    rownames(fit_L) <- sample.names
  }
  rownames(anno) <- sample.names
  cols <- colorRampPalette(c("gray96","red"))(50)
  brks <- seq(min(fit_L), max(fit_L),length.out= 50)
  rows <- order(anno$subpop)
  p <- pheatmap(fit_L[rows,], cluster_rows = FALSE, cluster_cols = FALSE,
          show_rownames = FALSE, annotation_row = anno, annotation_names_row = FALSE,
          annotation_colors = anno_colors, angle_col = 45, fontsize = 8, color = cols, breaks = brks, main = "")
  print(p)
}
```

```{r, eval=FALSE}
#plot heatmap
grouping_col <- rainbow(5)
names(grouping_col) <- levels(samples$subpop)
colnames(L.scaled) <- paste0('k',c(1:ncol(L.scaled)))
loadings_heatmap(fit_L = L.scaled, grouping = samples$subpop, sample.names = rownames(L.scaled), grouping_col)
```
  
From the heatmap, we see that almost all samples have membership in factor 1, supporting the interpretation of the first factor as an 'intercept factor'. We also see that the loadings of factors 2, 4, 5, and 6 generally correspond with NK cells, CD34+ cells, T cells, and B cells, respectively.  

Factors capture patterns of relative expression
--------------------
Each factor is represented as a vector of $m$ expression levels (one for each gene). All of the factors are stored in an $m \times K$ matrix. The entries of this matrix approximately represent the log-fold change (LFC) of a given gene in a given factor.  
    
For the scaled factor matrix, the entries can be interpreted as the log-fold-change of a given gene in a given factor if a sample has a membership level of 1 for that factor.    

```{r, eval=FALSE}
dim(fit.nn$F_pm)
```
    
As mentioned previously, the first factor is often interpreted as an intercept factor. It is usually loaded on all of the samples, and its expression levels across genes is relatively 'flat', i.e. there isn't a subset of genes that are more highly-expressed.    
    
Other factors may have (relatively) high levels of expression in only a subset of genes. In addition, genes may be highly expressed in only a subset of factors. For example, genes CD79A and CD79B are important to B cells, and thus we would expect them to have higher expression in factor 6.  

```{r, eval=FALSE}
genes <- pbmc_facs$genes
rbind(fit.nn$F_pm[genes$symbol == "CD79B",],
      fit.nn$F_pm[genes$symbol == "CD79A",])
```

Annotating topics by differentially expressed genes
--------------------
When interpreting the factors, we can look at the top driving genes for each factor. For example, for a given factor, we can find the genes with the top 10 highest expression levels.  

```{r, eval=FALSE}
get_top_genes <- function(fit, threshhold = 2, rowname_F){
  FF <- fit$F_pm
  LL <- fit$L_pm
  Fnorm <- t(t(FF)*apply(LL,2,max))
  rownames(Fnorm) <- rowname_F
  top_genes_list <- list()
  for (k in 1:ncol(Fnorm)){
    cur.genes <- rownames(Fnorm)[Fnorm[, k] > log(threshhold)]
    F.tmp <- Fnorm[cur.genes,]
    cur.genes <- rownames(F.tmp)[order(F.tmp[, k], decreasing = TRUE)] #re-order the genes
    lfc <- round(Fnorm[cur.genes, k], 3)
    dat <- data.frame(gene=cur.genes, lfc=lfc)
    rownames(dat) <- 1:nrow(dat)
    top_genes_list[[k]] <- dat # add the dataframe to the list of top driving genes
  }
  names(top_genes_list) <- paste0('Factor', c(1:ncol(Fnorm)))
  return(top_genes_list)
}
```
  
These are the top genes for factor 6:
```{r, eval=FALSE}
top_genes_list <- get_top_genes(fit.nn, rowname_F = genes$symbol)
print(top_genes_list[[6]][1:10,])
```
    
We can also visualize the top genes for a given factor with a volcano plot. A volcano plot is a type of scatter plot. On the x-axis, we plot the log-fold-change in expression levels for a given gene in a given factor. On the y-axis, we plot the mean expression level of the given gene across all cells in the (transformed) data matrix.

```{r, eval=FALSE}
volcano_plot <- function(gene, lfc, z, lfsr, k){
pdat <- data.frame(gene = gene, 
                   lfc  = lfc[,k], 
                   z    = abs(z), 
                   lfsr = lfsr[,k],
                   stringsAsFactors = FALSE)
pdat <- transform(pdat,lfsr = cut(lfsr, c(-1,0.001,0.01,0.05,Inf)))
rows  <- with(pdat, which(!(abs(lfc) > quantile(abs(lfc),0.999))))
pdat[rows, "gene"] <- ""
p <- ggplot(pdat, aes(x = lfc, y = z, color = lfsr, label = gene)) +
  geom_point() + 
  geom_text_repel(color = "black", size = 2.3, segment.color = "black",
                  segment.size = 0.25, min.segment.length = 0,
                  max.overlaps = Inf, na.rm = TRUE) +
  scale_color_manual(values = c("coral","orange","gold","deepskyblue")) +
  labs(x = "log-fold change", y = "mean expression level") + 
  theme_cowplot(font_size = 10)
print(p)
}
```
  
This is the volcano plot for factor 6:

```{r, eval=FALSE}
z <- (1/nrow(shifted_log_counts))*colSums(shifted_log_counts)
volcano_plot(genes$symbol, fit.nn$F_pm, as.matrix(z, ncol = 1), fit.nn$F_lfsr, 6)
```

We see that the top 10 most highly-expressed genes for factor 6 are identified in the volcano plot.

```{r session-info}
sessionInfo()
```

[zheng-2017]: https://doi.org/10.1038/ncomms14049
[fasttopics-vignette]: https://stephenslab.github.io/fastTopics/articles/single_cell_rnaseq_basic.html
